---
title: Exercises and Resources
number_sections: TRUE
author: "Eawag Summer School in Environmental Systems Analysis 2023"
output:
  html_document:
    number_sections: FALSE
    toc: TRUE
    toc_depth: 2
    css: ../eawag.css
---

<!-- setup Julia environmental and JS Code -->
```{r, child = "../header.Rmd"}
```
# General information


All participants have different backgrounds and interests. For this
reason we tend to provide more exercises than you can solve in this
week. Exercises that cover fundamental concepts are marked with â˜…. We
recommend that everyone has a look at them. Exercises that are more
advanced are marked with ðŸ—².



# Software {.tabset}

For most exercises we provide hints and solutions in **R** and
**Julia**. You can attempt the exercises in any other language you are
comfortable with, however, is some cases you may find it difficult to
find corresponding packages (in particular for [Exercises 3](exercise_3.html)).

**R** is a popular free software environment for statistical
computing, with an extensive range of packages for a extensive array
of statistical procedures, plotting options, data handling, and
more. See this
[overview](https://cran.r-project.org/web/views/). However, as an
interpreted language, R's performance may be slow, and deploying it in
a production environment can pose certain limitations.

**Julia** is a more recent, well designed language developed at MIT
aiming at scientific computation. It utilizes a highly efficient
just-in-time compilation for execution speeds comparable to those of C
or Fortran.  This makes Julia a great option for computationally
intensive tasks like Bayesian inference. Being a newer language than
R, you may find more rough edges in some libraries and less online
resources. The core language, however, is stable and very well
documented. Note, for the exercises in this course you will not benefit a
lot from the faster speed as all our example models are tiny. Julia
may even appear slower than R in interactive use due to the
compilation step, however, the faster execution speed of Julia makes
this a worthwhile trade-off for larger problems. Another distinct
advantage of Julia is the availability of Automatic Differentiation
that can help a lot with optimization and posterior sampling.


## R
### Installation

1. Install R from the  official [webpage](https://www.r-project.org/).

2. If you do not have a favorite text editor, we recommend to install [R Studio](https://posit.co/download/rstudio-desktop/), the most
    widely used IDE for R.

### Install the packages used for the course

Run this code once to install all required packages for this
course. Check careful for any error messages!
```r
## all packages that we need to install
pkgs <- c("mcmcensemble", "IDPmisc", "FME", "rootSolve",
          "adaptMCMC", "mvtnorm", "sensitivity", "deSolve")

## install packages if needed
for(pkg in pkgs){
    if(!(pkg %in% row.names(installed.packages()))){
        print(paste("-- Install: ", pkg))
        install.packages(pkg, dependencies = TRUE)
    }
}
```


### R Resources

- [A (very) short introduction to R](assets/A (very) short introduction to R.pdf){target="_blank"}


## Julia

### Installation

1. Download and install the latest stable Julia version from the  [
  official webpage](https://julialang.org/). Read the platform
  specific instructions carefully.

2. If you do not have a favorite text editor, install
     [Visual Studio Code](https://code.visualstudio.com/), and the Visual Studio Code [Julia
     Plugin](https://www.julia-vscode.org/) following this [instructions](https://www.julia-vscode.org/docs/dev/gettingstarted/#Installation-and-Configuration).


### Install packages used for the course

Julia has a build-in package manager that also handles environments
(for Python users, think `pip` and `virtualenv` combined), see the documentation [here](https://pkgdocs.julialang.org/v1/). The required packages used in the exercises
are defined in the files:

- [`Project.toml`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/src/Project.toml){target="_blank"}
  (defines the needed packages)
- [`Manifest.toml`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/src/Manifest.toml){target="_blank"} (defines all indirect dependencies and exact versions)

Save them at a convenient location and run
the following code to install and precompile all packaged. Note, the
precompilation may take several minutes!
```julia
import Pkg
Pkg.activate("path to *directory* containing Manifest.toml")
Pkg.instantiate()
Pkg.status()
```

Later activate the environment at the beginning of every script
(before `using` the packages) with:
```julia
import Pkg
Pkg.activate("path to *directory* containing Manifest.toml")

using ...
```

### Julia resources

- Official [Julia webpage](https://julialang.org/) with downloads and
  links to many other resources.

- Official [Julia documentation](https://docs.julialang.org/en/v1/), rather
  technical.

- A [brief introduction](https://learnxinyminutes.com/docs/julia/) for
programmers

- [Julia forum](https://discourse.julialang.org/), where you find
    many questions answered.

- [Noteworthy Differences from other
  Languages](https://docs.julialang.org/en/v1/manual/noteworthy-differences/)
  such as Matlab, Python, R.


# Example models and data

### Data

Download this [zip file](
https://github.com/scheidan/SiamSummerSchoolExercises/blob/main/data/exercise_data.zip?raw=true) and extract it to convenient location. It
contains all `csv` files needed in the exercises.

### Models {.tabset}

The example models are defined in this file. Save it in a convenient
location.

#### R
- [`models.r`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/models/models.r){target="_blank"}

You can load it with `source("path_to/models.r")` in your scripts.

#### Julia
- [`models.jl`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/models/models.jl){target="_blank"}

You can load it with `include("path_to/models.r")` in your scripts.


# Exercises

- [Exercises 0](exercise_0.html): review of probability
  calculus, basics of R and Julia

- [Exercises 1](exercise_1.html): deriving and implementing likelihood
  functions, sensitivity analysis

- [Exercises 2](exercise_2.html): Bayesian computations, MCMC

- [Exercises 3](exercise_3.html): Adaptive Bayesian computations, MCMC

#### Advances Topics

This exercises may use specialized software packages.

- [Exercises 4](exercise_4.html): Inference for Hierarchical models
  with STAN and JAGS

- [Exercises 5](exercise_5.html): Approximative Bayesian Computation
  (ABC)


# Solutions

- [Solutions 0](exercise_0_solution.html)

- [Solutions 1](exercise_1_solution.html)

- [Solutions 2](exercise_2_solution.html)

- [Solutions 3](exercise_3_solution.html)


# I've found a bug or typo! ðŸœ

If you find any typos, errors, confusing instructions, bad solutions,
or you have general suggestions how to
improve the exercises, please open an
[issue here](https://github.com/scheidan/SiamSummerSchoolExercises/issues)
(PRs are welcome too). Thanks a lot!

# Version info {.tabset}

Here you find the exact package versions used to compile the solutions:
<details>

## R
```{r, echo=FALSE}
sI <- sessionInfo()
print(sI, RNG = FALSE, locale = FALSE)
```

## Julia
```{julia, echo=FALSE}
using InteractiveUtils: versioninfo
versioninfo()
```

```{julia , echo=FALSE}
import Pkg
Pkg.activate("../.")
Pkg.status()
```
</details>

Built at `r format(Sys.time(), '%d %B, %Y')`.
