---
title: Exercises and Resources
number_sections: TRUE
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Eawag Summer School in Environmental Systems Analysis 2023"
output:
  html_document:
    number_sections: FALSE
    toc: TRUE
    toc_depth: 2
    css: ../eawag.css
---

<!-- setup Julia environmental and JS Code -->
```{r, child = "../header.Rmd"}
```
# General information


All participants have different backgrounds and interests. For this
reason we provide more exercises than you can solve in this
week. Exercises that cover fundamental concepts are marked with ★. We
recommend that everyone has a look at them. Exercises that are more
advanced are marked with 🗲.



# Software {.tabset}

For most exercises we provide the hints and solutions in **R** and
**Julia**.

**R** is a popular free software environment for statistical
computing, with an extensive range of packages for a extensive array
of statistical procedures, plotting options, data handling, and
more. See this
[overview](https://cran.r-project.org/web/views/). However, as an
interpreted language, R's performance may be slow, and deploying it in
a production environment can pose certain limitations.

**Julia** is a more recent, well designed language developed at MIT
aiming at scientific computation. It utilizes a highly efficient
just-in-time compilation for execution speeds comparable to those of C
or Fortran.  This makes Julia a great option for computationally
intensive tasks like Bayesian inference. Being a newer language than
R, you may find more rough edges in some libraries and less online
resources. The core language, however, is stable and very well
documented. Note, for the exercises in this course you will benefit a
lot from the faster speed as all our example models are tiny. Julia
may even appear slower than R in interactive use due to the
compilation step, however, the faster execution speed of Julia makes
this a worthwhile trade-off for larger problems.


## R
### Installation

1. Install R from the  official [webpage](https://www.r-project.org/).

2. If you do not have a favorite text editor, we recommend to install [R Studio](https://posit.co/download/rstudio-desktop/), the most
    widely used IDE for R.

### Install the packages used for the course
TODO

### R Resources

- [A (very) short introduction to R](assets/A (very) short introduction to R.pdf){target="_blank"}


## Julia

### Installation

1. Download and install the the latest stable Julia version form the  [
  official webpage](https://julialang.org/). Read the platform
  specific instructions carefully.

2. If you do not have a favorite text editor, we recommend to install
     [Visual Studio Code](https://code.visualstudio.com/).

3. Install the Visual Studio Code [Julia
     Plugin](https://www.julia-vscode.org/) following this [instructions](https://www.julia-vscode.org/docs/dev/gettingstarted/#Installation-and-Configuration).


### Install packages used for the course
TODO

### Julia resources


- Official [Julia webpage](https://julialang.org/) with downloads and
  links to many other resources.

- Official [Julia documentation](https://docs.julialang.org/en/v1/), rather
  technical.

- A [brief introduction](https://learnxinyminutes.com/docs/julia/) for
programmers

- [Julia forum](https://discourse.julialang.org/), where you find
    many questions answered.

- [Noteworthy Differences from other
  Languages](https://docs.julialang.org/en/v1/manual/noteworthy-differences/)
  such as Matlab, Python, R.


# Example models and data

### Data

Download this [`zip` file](
https://github.com/scheidan/SiamSummerSchoolExercises/blob/main/data/exercise_data.zip?raw=true) and extract it to convenient location. It
contains all `csv` files needed in the exercises.

### Models {.tabset}

The example models are defined in this file. Save it in a convenient
location.

#### R
- [`models.r`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/models/models.r){target="_blank"}

You can load it with `source("path_to/models.r")` in your scripts.

#### Julia
- [`models.jl`](https://raw.githubusercontent.com/scheidan/SiamSummerSchoolExercises/main/models/models.jl){target="_blank"}

You can load it with `include("path_to/models.r")` in your scripts.


# Exercises

- [Exercises 0](exercise_0.html): review of probability
  calculus, basics of R and Julia

- [Exercises 1](exercise_1.html): deriving and implementing likelihood
  functions, sensitivity analysis

- [Exercises 2](exercise_2.html): Bayesian computations, MCMC

- [Exercises 3](exercise_3.html): Adaptive Bayesian computations, MCMC

#### Advances Topics

This exercises may use specialized software packages.

- [Exercises 4](exercise_4.html): Inference for Hierarchical models
  with STAN and JAGS

- [Exercises 5](exercise_5.html): Approximative Bayesian Computation
  (ABC)


# Solutions

- [Solutions 0](exercise_0_solution.html)

- [Solutions 1](exercise_1_solution.html)

- [Solutions 2](exercise_2_solution.html)

- [Solutions 3](exercise_3_solution.html)


# I've found a bug or typo! 🐜

If you find any typos, errors, confusing instructions, bad solutions,
or you have general suggestions how to
improve the exercises, please open an
[issue here](https://github.com/scheidan/SiamSummerSchoolExercises/issues)
(PRs are welcome too). Thanks a lot!

# Version info {.tabset}

The solutions have been compiled with the follow versions:

## R
```{r, echo=FALSE}
sI <- sessionInfo()
print(sI, RNG = FALSE, locale = FALSE)
```

## Julia
```{julia, echo=FALSE}
using InteractiveUtils: versioninfo
versioninfo()
```

```{julia , echo=FALSE}
import Pkg
Pkg.activate("../.")
Pkg.status()
```
