## Solutions

### Analytical expression

For the model "Monod" with additive normal i.i.d. noise, constructing the likelihood follows exactly the same procedure as in the linear case (Exercise 1.1), only with a simple change in the deterministic part of the model. So again, our complete probabilistic model is given by the deterministic part plus a Gaussian noise term,

$$
Y_i = y_{i,det} + Z_i  , \quad \quad Z_i \sim \mathcal{N}(0,\sigma)
\;,
$$

but now the deterministic part is given by

$$
y_{det}(x,\boldsymbol{\theta}_{monod}) =
r(C,r_{max}, K) =  \frac{r_{max} C}{K+C}
\;.
\label{eqn:monodDet}
$$

We therefore again have three model parameters in total: $\boldsymbol{\theta} = \{ \boldsymbol{\theta}_{monod},\sigma\} = \{r_{max}, K,\sigma\}$.

The expression for the likelihood is then similar to that of the linear model:

$$
L(\boldsymbol{\theta}) = p(\mathbf{y}|\mathbf{C},r_{max}, K,\sigma) = \frac{1}{(2 \pi)^{n/2} \sigma^n } \exp\left( {-\frac{1}{2\sigma^2}} \sum_{i=1}^n (y_i - r(C_i,r_{max},K))^2 \right)
\;.
$$



### Likelihood evaluation {.tabset}


#### R
The implementation then looks like this:


```{r}
source("../../models/models.r") # load the deterministic model

loglikelihood.monod <- function(yobs, C, par){

    ## deterministic part:
    y.det <- model.monod(par, C)

    ## Calculate loglikelihood:
    return( sum(dnorm(yobs, mean=y.det, sd=par['sigma'], log=TRUE)) )
}


```

The parameters with higher likelihood are more likely to have generated the data. In our case, this should be the parameter set $\{r_{\mathrm{max}} = 5,\,K = 3,\,\sigma=0.2\}$.

Plotting the data together with the model output given those parameters looks ok too:
```{r}
## read data
dat <- read.table("../../data/model_monod_stoch.csv", header=TRUE)
C <- dat$C
r <- dat$r

## compare the two parameter vectors
par1 <- c(r_max = 5, K = 3, sigma=0.2)
par2 <- c(r_max = 10, K = 4, sigma=0.2)

loglikelihood.monod(r, C, par1)
loglikelihood.monod(r, C, par2)

## prediction
par <- c(r_max = 5, K = 3, sigma=0.2)
r.det <- model.monod(par, C)

## plot
plot(C, r)
lines(C, r.det, col=2)
```

#### Julia
Implementing the loglikelihood
```{julia, results = 'hide'}
using ComponentArrays
using Distributions
include("../../models/models.jl") # load the deterministic model

function loglikelihood_monod(yobs, C, par)
    ## deterministic part:
    ydet = model_monod(C, par)

    ## Calculate loglikelihood:
    return sum(logpdf.(Normal.(ydet, par.sigma), yobs))
end
```
Read the data
```{julia, results = 'hide'}
monod_data = CSV.read(joinpath("..", "..", "data", "model_monod_stoch.csv"),
             DataFrame)

C = monod_data.C
yobs = monod_data.r
```
We can compute
```{julia}
## compare the two parameter vectors
par1 = ComponentVector(r_max = 5, K = 3, sigma=0.2)
par2 = ComponentVector(r_max = 10, K = 4, sigma=0.2)

loglikelihood_monod(yobs, C, par1)
loglikelihood_monod(yobs, C, par2)

```
Plotting
```{julia, results = 'hide'}
# use the better parameters to compute the deterministic predictions
y_det = model_monod(C, par1)
```

```{julia, out.width = '80%'}
scatter(C, yobs,
        labels = false,
        xlabel = "C",
        ylabel = "r");
plot!(C, y_det,
      labels = false)
```

#### Python
The implementation then looks like this:


```{python}

def loglikelihood_monod(yobs, C, par):
    """
    Calculate log-likelihood of observations yobs given parameters par and substrate concentration C.
    
    Arguments:
    ----------
    - yobs: array-like
        Observed data.
    - C: array-like
        Substrate concentration.
    - par: dictionary
        Parameters including:
        - 'r_max': maximum growth rate
        - 'K': half-saturation concentration
        - 'sigma': standard deviation of the normal distribution for the likelihood
        
    Returns:
    -------
    log_likelihood: float
        The calculated log-likelihood of the observed data.
    """
    
    # Deterministic part: calculate y.det using model_monod function
    y_det = model_monod(par, C)
    sigma = par[-1]
    # Calculate log-likelihood: sum of log probabilities of yobs given y_det and sigma
    log_likelihood = np.sum(norm.logpdf(yobs, loc=y_det, scale=sigma))
    
    return log_likelihood
```

The parameters with higher likelihood are more likely to have generated the data. In our case, this should be the parameter set $\{r_{\mathrm{max}} = 5,\,K = 3,\,\sigma=0.2\}$.

Plotting the data together with the model output given those parameters looks ok too:
```{python}
# Loading data
# Specify the path to the CSV file
file_path = r"../../data/model_monod_stoch.csv"

# Load the CSV file into a pandas DataFrame
dat = pd.read_csv(file_path, sep=" ")

C = dat['C']
r = dat['r']

# Define the two parameter vectors
# 'r_max': 5, 'K': 3, 'sigma': 0.2
# 'r_max': 10, 'K': 4, 'sigma': 0.2

par1 = [5, 3, 0.2]
par2 = [10, 4, 0.2]

# Define the two parameter vectors
#par1 = {'r_max': 5, 'K': 3, 'sigma': 0.2}
#par2 = {'r_max': 10, 'K': 4, 'sigma': 0.2}

# Calculate the log-likelihood using the loglikelihood_monod function
loglikelihood_par1 = loglikelihood_monod(r, C, par1)
loglikelihood_par2 = loglikelihood_monod(r, C, par2)

# Output the log-likelihood results
print("Log-likelihood for par1:", loglikelihood_par1)
print("Log-likelihood for par2:", loglikelihood_par2)
```

```{python}
# Assume C and r are already imported and available
# par is a vector with parameters 'r_max', 'K' and 'sigma'
par = [5, 3, 0.2]

# Predict the deterministic growth rate using model_monod function
r_det = model_monod(par, C)

# Plot the observed data
plt.figure(figsize=(14, 6))
sns.scatterplot(x=C, y=r, label='Observed data')

# Plot the deterministic growth rate as a line
sns.lineplot(x=C, y=r_det, color='red', label='Deterministic model')

# Add labels and a legend
plt.xlabel('Substrate concentration (C)')
plt.ylabel('Growth rate (r)')
plt.legend()

# Adjust layout and display the plots
plt.tight_layout()
plt.show()
```

