# 4. Solutions

## Analytical expression for the likelihood

The probability for an individual with mortality rate $\lambda$ to survive until time $t$ is given by the exponential
$$
S(t,\lambda)=e^{-\lambda t}\,.
$$

Thus, the probability of the individual to die within time interval $[t_{i-1},t_i]$ is given by
$$
p_{i,\lambda}=S(t_{i-1},\lambda) - S(t_i,\lambda)\,.
$$

Assume that we have $N$ individuals, each with the same mortality rate.
We want to calculate the probability that $y_1$ individuals die in time interval $[t_0,t_1]$, $y_2$ individuals die in time interval $[t_2,t_3]$ etc.
Those individuals who survive the whole experiment die in the time interval $[t_n,\infty]$, and this happens with probability $p_{n+1}=S(t_n)$.
For the time being, assume that we distinguish the individuals and want to calculate the probability that individual 1 dies in, say, time interval $[t_3,t_4]$, individual 2 in time interval, say, $[t_1,t_2]$ etc. in such a way that the counts $\bf y$ are assumed.
Due to the independence of the individuals, the probability, for such an event, is then simply given by the product
$$
p_{1,\lambda}^{y_1} \cdot p_{2,\lambda}^{y_2} \cdot \ldots \cdot \, p_{n,\lambda}^{y_n}\,.
$$

Since we do not distinguish individuals, we have to multiply this product by a so-called multinomial coefficient, which counts the number of ways to put $N$ individuals into $n+1$ buckets, with $y_1$ ending up in the first bucket etc.
Thus, the likelihood function, for model parameter $\lambda$, given an output $\bf y$ of our survival model is described by the so-called multinomial distribution
$$
L_{\text{survival}}(\lambda,{\bf y})
=
\frac{N!}{y_1!\cdots y_{n+1}!} p_{1,\lambda}^{y_1} \cdot p_{2,\lambda}^{y_2} \cdot \ldots \cdot p_{n+1,\lambda}^{y_{n+1}}\,.
$$


## Forward simulation {.tabset}

### R
The simplest way to simulate a model output is to use the R function `rmultinom`.
Alternatively, we can simulate the death time, for each individual, drawing from the exponential distribution
$$
f(t|\lambda)=\lambda e^{-\lambda t}\,,
$$
which is the negative derivative of the survival probability $S(t)$.
After that, we count how many individuals die within the different time intervals.

```{r}
## parameters for model "Survival":

N            <- 30                 # Number of individuals
t            <- 1:5                # Times at which the numbers of deaths are counted
par.survival <- c(lambda=0.2)    # Parameter vector of the model "Survival"


## Simulation from  model "Survival".
## Here we  sample from the multinomial distribution implemented in R:
model.survival <- function(N, t, par){
    ## Calculate survival probabilities at measurement points
    ## and add zero "at time-point infinity":
    S <- exp(-par*c(0,t))
    S <- c(S,0)

    ## Calcute probabilities to die within observation windows:
    p <- -diff(S)

    ## Simulate from multinomial distribution:
    y <- rmultinom(1, N, p)[-length(p) ]

    return(y)
}


## Alternatively, we generate the exponentially distributed
## times of death and bin them
model.survival.alt <- function(N, t, par) {
    ## Generate deaths times:
    T <- rexp(N, par)

    ## Count number of deaths within observation windows:
    t2 <- c(t,6)     # construct the bins
    y <- table(cut(T+1, breaks = t2)) # do the data binning

    return(y)
}


## Generate n.sample simulations from the probabilistic model
n.sample  <- 1000
pred.survival <- matrix(nrow=n.sample, ncol=length(t))
for (i in 1:n.sample){
	pred.survival[i,] <- model.survival(N, t, par.survival)
}


## Make a boxplot:
boxplot(pred.survival, xlab="time", ylab="# death")
```

### Julia
todo


## Likelihood evaluation {.tabset}

### R

```{r}

loglikelihood.survival <- function(y, N, t, par){

    ## Calculate survival probabilities at measurement points
    ## and add zero "at time-point infinity":
    S <- exp(-par*c(0,t))
    S <- c(S,0)

    ## Calcute probabilities to die within observation windows:
    p <- -diff(S)

    ## Add number of survivors at the end of the experiment:
    y <- c(y, N-sum(y))

    ## Calculate log-likelihood of multinomial distribution at point y:
    LL <- dmultinom(y, prob=p, log=TRUE)

    return(LL)
}

## simulate data
y.sim <- model.survival(N,t,par.survival)

## evaluation likelihood
loglikelihood.survival(y.sim,N,t,par.survival)
```

### Julia
TODO
