# 3. Solutions

## Deterministic model simulation {.tabset}

### R
#### Monod Model
```{r }
source("../../models/models.r") # load the model

## parameters
C.monod <- seq(0, 10, 0.1)    # concentrations at which model output should be calculated
par.monod <- c(r_max=5, K=3)   # define parameter vector for the model "Monod"

## run model
r.det <- model.monod(par.monod, C.monod)

## plot result
plot(C.monod, r.det, col=2, type="l", ylab="r")
```

#### Growth Model
```{r }
source("../../models/models.r") # load the model

## parameters
par.growth <- c(mu=4,
                K=10,
                b=1,
                Y=0.6,
                C_M_ini=10,
                C_S_ini=50)

times <- seq(0, 2, length=101) # time resolution of solution

## run model
out <- model.growth(par.growth, times)

## plot results
plot(out$time, out$C_M, col=3, type="l", ylim=c(0, 51),
     xlab="time", ylab="concentration")
lines(out$time, out$C_S, col=4)
legend("topright", legend=c("microorganisms", "substrate"),
       col=c(3,4), lty=1)
```

### Julia
TODO

## Stochastic model simulation  {.tabset}

The growth model takes the input variables  $\mathbf{C} =
(C_1,...,C_{n})$ and returns the deterministic growth rate $\mathbf{r}_{\mathrm{det}} = \mathbf{r}(\mathbf{C},r_{\mathrm{max}},K) = \{r(C_1,r_{\mathrm{max}},K),\dots,r(C_n,r_{\mathrm{max}},K)\}$.
We simulate the measurement errors by generating $n$ random
normal values with mean $0$ and standard deviation $\sigma$ and add
them to the deterministic model output.

### R
```{r }
simulate.monod.stoch <- function(par, C){
    ## run model
    r.det <- model.monod(par, C)

    ## generate noise
    z <- rnorm(length(C.monod), 0, par["sigma"])

    return(r.det + z)
}

## parameters
C.monod <- seq(0, 10, 0.1)    # concentrations at which model output should be calculated
par.monod <- c(r_max=5, K=3, sigma=0.2)   # define parameter vector for the model "Monod"
n.rep <- 1000

## run `n.rep` simulations
r.sims <- replicate(n=n.rep, simulate.monod.stoch(par.monod, C.monod))
dim(r.sims)

## compute quantiles
quant <- apply(r.sims, MARGIN=1, FUN=quantile, probs=c(0.1, 0.9))
dim(quant)

## plot result
plot(C.monod, r.det, type="n", ylab="r")
polygon(c(C.monod,rev(C.monod)), c(quant[1,],rev(quant[2,])), col = "grey85")
lines(C.monod, r.det, col=2, lwd=2, ylab="r")
```

### Julia
TODO
