## Solutions

### Analytical expression for the likelihood

The probability for an individual with mortality rate $\lambda$ to survive until time $t$ is given by the exponential
$$
S(t,\lambda)=e^{-\lambda t}\,.
$$

Thus, the probability of the individual to die within time interval $[t_{i-1},t_i]$ is given by
$$
p_{i,\lambda}=S(t_{i-1},\lambda) - S(t_i,\lambda)\,.
$$

Assume that we have $N$ individuals, each with the same mortality rate.
We want to calculate the probability that $y_1$ individuals die in time interval $[t_0,t_1]$, $y_2$ individuals die in time interval $[t_2,t_3]$ etc.
Those individuals who survive the whole experiment die in the time interval $[t_n,\infty]$, and this happens with probability $p_{n+1}=S(t_n)$.
For the time being, assume that we distinguish the individuals and want to calculate the probability that individual 1 dies in, say, time interval $[t_3,t_4]$, individual 2 in time interval, say, $[t_1,t_2]$ etc. in such a way that the counts $\bf y$ are assumed.
Due to the independence of the individuals, the probability, for such an event, is then simply given by the product
$$
p_{1,\lambda}^{y_1} \cdot p_{2,\lambda}^{y_2} \cdot \ldots \cdot \, p_{n,\lambda}^{y_n}\,.
$$

Since we do not distinguish individuals, we have to multiply this product by a so-called multinomial coefficient, which counts the number of ways to put $N$ individuals into $n+1$ buckets, with $y_1$ ending up in the first bucket etc.
Thus, the likelihood function, for model parameter $\lambda$, given an output $\bf y$ of our survival model is described by the so-called multinomial distribution
$$
p({\bf y} \mid \lambda) = L_{\text{survival}}(\lambda)
=
\frac{N!}{y_1!\cdots y_{n+1}!} p_{1,\lambda}^{y_1} \cdot p_{2,\lambda}^{y_2} \cdot \ldots \cdot p_{n+1,\lambda}^{y_{n+1}}\,.
$$


### Forward simulation {.tabset}

#### R
The simplest way to simulate a model output is to use the R function `rmultinom`.
Alternatively, we can simulate the death time, for each individual, drawing from the exponential distribution
$$
f(t|\lambda)=\lambda e^{-\lambda t}\,,
$$
which is the negative derivative of the survival probability $S(t)$.
After that, we count how many individuals die within the different time intervals.

```{r}
## parameters for model "Survival":

N            <- 30                 # Number of individuals
t            <- 1:5                # Times at which the numbers of deaths are counted
par.survival <- c(lambda=0.2)    # Parameter vector of the model "Survival"


## Simulation from  model "Survival".
## Here we  sample from the multinomial distribution implemented in R:
model.survival <- function(N, t, par){
    ## Calculate survival probabilities at measurement points
    ## and add zero "at time-point infinity":
    S <- exp(-par*c(0,t))
    S <- c(S,0)

    ## Calcute probabilities to die within observation windows:
    p <- -diff(S)

    ## Simulate from multinomial distribution:
    y <- rmultinom(1, N, p)[-length(p)]

    return(y)
}


## Alternatively, we generate the exponentially distributed
## times of death and bin them
model.survival.alt <- function(N, t, par) {
    ## Generate deaths times:
    T <- rexp(N, par)

    ## Count number of deaths within observation windows:
    t2 <- c(t,6)     # construct the bins
    y <- table(cut(T+1, breaks = t2)) # do the data binning

    return(y)
}


## Generate n.sample simulations from the probabilistic model
n.sample  <- 1000
pred.survival <- matrix(nrow=n.sample, ncol=length(t))
for (i in 1:n.sample){
	pred.survival[i,] <- model.survival(N, t, par.survival)
}


## Make a boxplot:
boxplot(pred.survival, xlab="time", ylab="# death")
```

#### Julia
Create a function survival, whose arguments are:

* `N,` the number of individuals,

* `t`, a vector of the form 1:k

* `λ`, the mortality rate.

Returns a random vector `y` of number of deaths per bin.
```{julia, results = 'hide'}
function survival(N::Int64, t, λ)
    # S is a vector whose first component is 1 and last is 0,
    # because the probability to be alive at time t=0 is 1,
    # and the probability to be alive at time t=infinity is 0
    S = exp.(-λ .* t)
    S = [1; S; 0]
    p = -diff(S)

    y = rand(Multinomial(N, p))[1:(end-1)]
    return y
end

# setting parameter
N = 30
t = 1:5
λ = 0.2
n_sample = 1000

pred_survival = Matrix{Int}(undef, n_sample, length(t))
for i in 1:n_sample
    pred_survival[i,:] = survival(N, t, λ)
end
```
Plotting
```{julia, out.width = '80%'}
using StatsPlots

boxplot(pred_survival,
        labels = false, xlab="time", ylab="# deaths")

```

#### Python
Create a function survival, whose arguments are:

* `N,` the number of individuals,

* `t`, a vector of the form 1:k

* `λ`, the mortality rate.

Returns a random vector `y` of number of deaths per bin.

```{python}
# Parameters for the survival model
N = 30  # Number of individuals
t = np.array([1, 2, 3, 4, 5])  # Times at which the numbers of deaths are counted
par_survival = [0.2]  # Parameter vector of the model "Survival": 'lambda'

# Function for the survival model
def model_survival(N, t, par):
    lambda_ = par[0]
    
    # Calculate survival probabilities at measurement points and add zero "at time-point infinity"
    S = np.exp(-lambda_ * np.concatenate(([0], t)))
    S = np.append(S, 0)  # Add zero "at time-point infinity"
    
    # Calculate probabilities to die within observation windows
    p = -np.diff(S)
    
    # Simulate from multinomial distribution
    y = multinomial.rvs(N, p)[:-1]
    
    return y

# Generate simulations from the probabilistic model
n_sample = 1000
pred_survival = np.zeros((n_sample, len(t)))

for i in range(n_sample):
    pred_survival[i, :] = model_survival(N, t, par_survival)
    
# Convert the NumPy array to a pandas DataFrame (as input to seaborn)
pred_survival_df = pd.DataFrame(pred_survival, columns=t)

# Plotting a boxplot
plt.figure(figsize=(8, 6))
sns.boxplot(data=pred_survival_df)
plt.xlabel("Time")
plt.ylabel("# Deaths")
plt.title("Boxplot of # Deaths at Different Times")

# Adjust layout and display the plots
plt.tight_layout()
plt.show()
```

### Likelihood evaluation {.tabset}

#### R

```{r}

loglikelihood.survival <- function(y, N, t, par){

    ## Calculate survival probabilities at measurement points
    ## and add zero "at time-point infinity":
    S <- exp(-par*c(0,t))
    S <- c(S,0)

    ## Calcute probabilities to die within observation windows:
    p <- -diff(S)

    ## Add number of survivors at the end of the experiment:
    y <- c(y, N-sum(y))

    ## Calculate log-likelihood of multinomial distribution at point y:
    LL <- dmultinom(y, prob=p, log=TRUE)

    return(LL)
}

## simulate data
y.sim <- model.survival(N, t, par.survival)

## evaluation likelihood
loglikelihood.survival(y.sim, N, t, par.survival)
```

#### Julia
```{julia, results = 'hide'}
function loglikelihood_survival(N::Int, t, y, λ)
    S = exp.(-λ .* t)
    S = [1; S; 0]
    p = -diff(S)
    ## Add number of survivors at the end of the experiment:
    push!(y, N-sum(y))

    logpdf(Multinomial(N, p), y)
end
```
```{julia}
# Generating and checking
N = 30
t = 1:5
λ = 0.2
y = survival(N, t, λ)

loglikelihood_survival(N, t, y, λ)
```

```{python}
def loglikelihood_survival(y, N, t, par):
    """
    Calculate the log-likelihood of the observed data y given the parameters par
    and the times t using the multinomial distribution.

    Arguments:
    - y: observed data (number of deaths at each time point).
    - N: total number of individuals in the experiment.
    - t: time points at which the number of deaths are counted.
    - par: dictionary containing the lambda parameter for the model.

    Returns:
    - LL: log-likelihood value.
    """
    lambda_ = par[0]
    
    # Calculate survival probabilities at measurement points and add zero "at time-point infinity"
    S = np.exp(-lambda_ * np.concatenate(([0], t)))
    S = np.append(S, 0)
    
    # Calculate probabilities to die within observation windows
    p = -np.diff(S)
    
    # Add number of survivors at the end of the experiment
    y = np.append(y, N - np.sum(y))
    
    # Calculate log-likelihood of multinomial distribution at point y
    LL = multinomial.logpmf(y, N, p)
    
    return LL

# Simulate data using the model_survival function
def model_survival(N, t, par):
    lambda_ = par[0]
    
    # Calculate survival probabilities at measurement points and add zero "at time-point infinity"
    S = np.exp(-lambda_ * np.concatenate(([0], t)))
    S = np.append(S, 0)
    
    # Calculate probabilities to die within observation windows
    p = -np.diff(S)
    
    # Simulate from multinomial distribution
    y = multinomial.rvs(N, p)[:-1]
    
    return y

# Parameters for the survival model
N = 30  # Number of individuals
t = np.array([1, 2, 3, 4, 5])  # Times at which the numbers of deaths are counted
par_survival = [0.2]  # Parameter vector of the model "Survival"

# Set seed for reproducibility
np.random.seed(1)

# Simulate data
y_sim = model_survival(N, t, par_survival)

# Evaluate log-likelihood
LL = loglikelihood_survival(y_sim, N, t, par_survival)
print(f"Log-likelihood: {LL}")
```

