## Solutions

### Deterministic model simulation {.tabset}

#### R
##### Monod Model
```{r }
source("../../models/models.r") # load the model

## parameters
C.monod <- seq(0, 10, 0.1)    # concentrations at which model output should be calculated
par.monod <- c(r_max=5, K=3)   # define parameter vector for the model "Monod"

## run model
r.det <- model.monod(par.monod, C.monod)

## plot result
plot(C.monod, r.det, col=2, type="l", ylab="r")
```

##### Growth Model
```{r }
source("../../models/models.r") # load the model

## parameters
par.growth <- c(mu=4,
                K=10,
                b=1,
                Y=0.6,
                C_M_ini=10,
                C_S_ini=50)

times <- seq(0, 2, length=101) # time resolution of solution

## run model
out <- model.growth(par.growth, times)

## plot results
plot(out$time, out$C_M, col=3, type="l", ylim=c(0, 51),
     xlab="time", ylab="concentration")
lines(out$time, out$C_S, col=4)
legend("topright", legend=c("microorganisms", "substrate"),
       col=c(3,4), lty=1)
```

#### Julia
##### Monod model
```{julia, results = 'hide'}
using ComponentArrays
include("../../models/models.jl") # load the model

#set parameters
par = ComponentVector(r_max = 5, K = 3)
C_monod = 0:0.1:10

# run deterministic model monod
Y_monod = model_monod(C_monod, par)
```
Plotting
```{julia, out.width = '80%'}
plot(C_monod, Y_monod,
    labels = false,
    xlabel = "C",
    ylabel = "r")
```



##### Growth model
```{julia, results = 'hide'}
using ComponentArrays
include("../../models/models.jl") # load the model

# set parameters
par = ComponentVector(mu=4, K=10, b=1, Y=0.6,
                      inits = (C_M=10.0, C_S=50.0))
times = 0:0.01:3

# run model
res = model_growth(times, par)

```
Plotting
```{julia, out.width = '80%'}
plot(res.time, res.C_M,
     label = "C_M",
     xlabel = "time",
     ylabel = "concentrations",
     title = "Deterministic model growth");
plot!(res.time, res.C_S, label="C_S")

```

#### Python
##### Monod Model
```{python}
# Parameters
C_monod = np.arange(0, 10.1, 0.1)  # concentrations at which model output should be calculated
par_monod = [5, 3]   # define parameter vector for the model "Monod"

# Run model
r_det = model_monod(par_monod, C_monod)

# Plot result
plt.figure(figsize=(14, 6))
sns.lineplot(x=C_monod, y=r_det, color='red')
plt.ylabel('r')
plt.xlabel('C')
plt.title('Model Monod output')

# Adjust layout and display the plots
plt.tight_layout()
plt.show()
```

##### Growth Model
```{python}
# Parameters
# 'mu', 'K', 'b','Y','C_M_ini','C_S_ini'

par_growth = [4, 10, 1, 0.6, 10, 50]

times = np.linspace(0, 2, 101)  # time resolution of solution

# Run model
out = model_growth(par_growth, times)

# Plot results
plt.figure(figsize=(14, 6))
sns.lineplot(x=out['time'], y=out['C_M'], color='green', label='microorganisms')
sns.lineplot(x=out['time'], y=out['C_S'], color='blue', label='substrate')
plt.ylim(0, 51)
plt.xlabel('time')
plt.ylabel('concentration')
plt.legend(loc='upper right')
plt.title('Growth of microorganisms and substrate over time')
plt.show()
```


### Stochastic model simulation  {.tabset}

The growth model takes the input variables  $\mathbf{C} =
(C_1,...,C_{n})$ and returns the deterministic growth rate $\mathbf{r}_{\mathrm{det}} = \mathbf{r}(\mathbf{C},r_{\mathrm{max}},K) = \{r(C_1,r_{\mathrm{max}},K),\dots,r(C_n,r_{\mathrm{max}},K)\}$.
We simulate the measurement errors by generating $n$ random
normal values with mean $0$ and standard deviation $\sigma$ and add
them to the deterministic model output.

#### R
```{r}
source("../../models/models.r") # load the model

simulate.monod.stoch <- function(par, C){
    ## run model
    r.det <- model.monod(par, C)

    ## generate noise
    z <- rnorm(length(C), 0, par["sigma"])

    return(r.det + z)
}

## parameters
C.monod <- seq(0, 10, 0.1)    # concentrations at which model output should be calculated
par.monod <- c(r_max=5, K=3, sigma=0.2)   # define parameter vector for the model "Monod"
n.rep <- 1000

## run `n.rep` simulations
r.sims <- replicate(n=n.rep, simulate.monod.stoch(par.monod, C.monod))
dim(r.sims)

## compute quantiles
quant <- apply(r.sims, MARGIN=1, FUN=quantile, probs=c(0.1, 0.9))
dim(quant)

## plot result
plot(C.monod, r.det, type="n", ylab="r")
polygon(c(C.monod,rev(C.monod)), c(quant[1,],rev(quant[2,])), col = "grey85")
lines(C.monod, r.det, col=2, lwd=2, ylab="r")
```

#### Julia
```{julia, results = 'hide'}
using ComponentArrays
include("../../models/models.jl") # load the deterministic model

# function to simulate stochastic realisations
function simulate_monod_stoch(C, par)
    Ydet = model_monod(C, par)
     z = rand(Normal(0, par.sigma), length(Ydet)) # adding noise
    Ydet .+ z
end

# parameters
par = ComponentVector(r_max = 5, K = 3, sigma=0.2)
C_monod = 0:0.1:10
n_sample = 1000

monod_stoch = hcat((simulate_monod_stoch(C_monod, par) for _ in 1:n_sample)...)

#compute quantile
low_quantile = [quantile(monod_stoch[i,:], 0.025) for i in 1:length(C_monod)]
med_quantile = [quantile(monod_stoch[i,:], 0.5) for i in 1:length(C_monod)]
upper_quantile = [quantile(monod_stoch[i,:], 0.975) for i in 1:length(C_monod)]
```
Plotting
```{julia, out.width = '80%'}
plot(C_monod, upper_quantile,
    fillrange = low_quantile,
    labels = false,
    xlabel = "C",
    ylabel = "r");
plot!(C_monod, med_quantile,
    labels = false)
```

#### Python
```{python}
def simulate_monod_stoch(par, C):
    """
    Simulate the Monod model with stochastic noise.
    
    Arguments:
    ----------
    - par: Array containing the following parameters:
        - r_max: maximum growth rate
        - K: half-saturation concentration
        - sigma: standard deviation of noise
    - C: numpy array containing substrate concentrations
    
    Value:
    ------
    A numpy array representing the growth rate with stochastic noise added.
    """
    # Run deterministic model
    r_det = model_monod(par, C)

    # Generate noise using a normal distribution with mean 0 and standard deviation `sigma`
    sigma = par[-1]
    z = np.random.normal(0, sigma, size=len(C))
    
    # Add noise to the deterministic model results
    return r_det + z

# Parameters
C_monod = np.arange(0, 10.1, 0.1)  # concentrations at which model output should be calculated
par_monod = [5,  3,  0.2]  # define parameter array for the model "Monod"
n_rep = 1000  # number of simulations

# Run `n_rep` simulations
r_sims = np.array([simulate_monod_stoch(par_monod, C_monod) for _ in range(n_rep)])

# The dimensions of `r_sims` array will be (n_rep, len(C_monod))
print(r_sims.shape)  # Should print (1000, 101)

# Compute quantiles
# r_sims is a numpy array where each row is a simulation and each column represents a concentration
quant = np.quantile(r_sims, q=[0.1, 0.9], axis=0)

# The resulting `quant` is a 2D array where the first row contains the 10th percentile,
# and the second row contains the 90th percentile for each concentration.
print(quant.shape)  # Prints the shape of the resulting quantile array

# Define variables for plotting
C_monod = np.linspace(0, 10, 101)  # Concentrations at which model output is calculated

# Plot result
plt.figure(figsize=(14, 6))

# Plot deterministic model output using seaborn
sns.lineplot(x=C_monod, y=r_det, color='red', label='Deterministic model', linewidth=2)

# Create a shaded region between the quantiles (0.1 and 0.9)
plt.fill_between(C_monod, quant[0, :], quant[1, :], color='gray', alpha=0.5, label='10th-90th percentile')

# Set labels for the x and y axes
plt.xlabel('C')
plt.ylabel('r')

# Add a legend
plt.legend()

# Adjust layout and display the plots
plt.tight_layout()
plt.show()
```