## Solution {.tabset}

### R

```{r}
library(mcmcensemble)
```

For each walker (chain) an initial starting point must be defined. In
general, it is better to choose it in a region with high density. We could use an
optimizer to fine the mode, but here we just use a rather wide coverage.
```{r}
n.walkers <- 20
par.inits <- data.frame(r.max = runif(n.walkers, 1, 10),
                        K     = runif(n.walkers, 0, 5),
                        sigma = runif(n.walkers, 0.05, 2))

```

```{r}
EMCEE <- MCMCEnsemble(
    f           = logposterior.monod,
    inits       = par.inits,
    max.iter    = 10000,
    n.walkers   = n.walkers,
    method      = "stretch",
    coda        = TRUE
)
```

```{r}
plot(EMCEE$samples)
```

Note, the more walkers (chains) we have, the shorter the chains. This means we have to "pay" the burn in for every single
chain. Therefore, going too extreme with the number of chains is not beneficial.
```{r}
n.walkers <- 1000
par.inits <- data.frame(r.max = runif(n.walkers, 1, 10),
                        K     = runif(n.walkers, 0, 5),
                        sigma = runif(n.walkers, 0.05, 2))

EMCEE <- MCMCEnsemble(
    f           = logposterior.monod,
    inits       = par.inits,
    max.iter    = 10000,
    n.walkers   = n.walkers,
    method      = "stretch",
    coda        = TRUE
)

plot(EMCEE$samples)
```

### Julia

```{julia, out.width = '80%'}
using ComponentArrays
using KissMCMC: emcee
using Plots
using StatsPlots

# number of walkers (parallel chains)
n_walkers = 10;

## We need a vector of inital values, one for each walkers.
θinit = ComponentVector(r_max = 2.5, K=1.4, sigma=0.25); # prior mean
## We add some randomnesses to make sure that they do not start
## from the same point.
θinits = [θinit .* rand(Normal(0, 0.1), 3) for _ in 1:n_walkers];

# Run sampler
samples, acceptance_rate, lp = emcee(logposterior_monod,
                                     θinits;
                                     niter = 10_000, # total number of density evaluations
                                     nburnin = 0);

# Converting into `MCMCChains.Chains` object for plotting.
X = permutedims(
    cat((hcat(samples[i]...) for i in 1:n_walkers)..., dims=3),
    [2, 1, 3]);
chn = Chains(X, labels(θinit))
```
Note, that our chains are only of length 1000. So we have a lot of
computation used for the burn-in phase.
```{julia, out.width = '80%'}
# plotting
plot(chn)
corner(chn)
# removing burn-in:
corner(chn[250:end,:,:])
```
