## Solution {.tabset}

### R

```{r, results="hide"}
library(adaptMCMC)
library(IDPmisc)
```
The mean of the prior seems to be a reasonable point to start the
sampler. Alternatively, we could try to find the point of the maximum
posterior density with an optimizer. For the covariance matrix of
the jump distribution we use the standard deviation of the prior and assume independence.

```{r}
par.init <- prior.monod.mean
jump.cov <- diag(prior.monod.sd/2)

RAM <- adaptMCMC::MCMC(p        = logposterior.monod,
                       n        = 10000,
                       init     = par.init,
                       scale    = jump.cov,
                       adapt    = TRUE,
                       acc.rate = 0.5,
                       showProgressBar = FALSE)
```
The acceptance rate is matched closely:
```{r}
RAM$acceptance.rate
```

The adapted covariance matrix has a lot of correlation between
$r_{max}$ and $K$:
```{r}
RAM$cov.jump
cov2cor(RAM$cov.jump) # rescale as correlation matrix
```

```{r}

## plot chains
samp.coda <- convert.to.coda(RAM)
plot(samp.coda)


## 2d marginals
IDPmisc::ipairs(RAM$samples) # prettier versions of pairs()
```

### Julia

```{julia, out.width = '80%'}
using ComponentArrays
using AdaptiveMCMC
using MCMCChains
using Plots
using StatsPlots

θinit = ComponentVector(r_max = 2.5, K=1.4, sigma=0.25) # use prior mean
res = adaptive_rwm(θinit, logposterior_monod,
                   10_000;
                   algorithm = :ram,
                   b=1,   # length of burn-in. Set to 1 for no burn-in.
                   );
# convert to MCMCChains for summary and plotting
chn = Chains(res.X', labels(θinit))
plot(chn)
corner(chn)
```
