## Solution {.tabset}

### R

```{r}
library(mcmcensemble)
```

For each walker (chain) a initial starting point must be defined. We
make sure that the walkers are spread out wide and cover the
high density posterior region.
```{r}
n.walkers <- 20
par.inits <- data.frame(r.max = runif(n.walkers, 1, 10),
                        K     = runif(n.walkers, 0, 5),
                        sigma = runif(n.walkers, 0.05, 2))

```

```{r}
EMCEE <- MCMCEnsemble(
    f           = logposterior.monod,
    inits       = par.inits,
    max.iter    = 10000,
    n.walkers   = n.walkers,
    method      = "stretch",
    coda        = TRUE
)
```

```{r}
plot(EMCEE$samples)
```

Note, the more walkers (chains) we have, the shorter the chains. This means we have to "pay" the burn in for every single
chain. Therefore, going too extreme with the number of chains is not beneficial.
```{r}
n.walkers <- 1000
par.inits <- data.frame(r.max = runif(n.walkers, 1, 10),
                        K     = runif(n.walkers, 0, 5),
                        sigma = runif(n.walkers, 0.05, 2))

EMCEE <- MCMCEnsemble(
    f           = logposterior.monod,
    inits       = par.inits,
    max.iter    = 10000,
    n.walkers   = n.walkers,
    method      = "stretch",
    coda        = TRUE
)

plot(EMCEE$samples)
```

### Julia

TODO
