---
title: Rmarkdown Template for Summer School Exercises
number_sections: TRUE
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Eawag Summer School in Environmental Systems Analysis"
output:
  html_document:
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
params:
  showsolutions: TRUE
---

# Setup R, Julia, and Python environments {-}

### R  {-}
```{r, echo=TRUE}
library(renv, warn.conflicts = FALSE)
renv::use(lockfile="renv.lock")
.libPaths()  # package load paths
```

### Julia  {-}
```{julia, echo=TRUE}
import Pkg
Pkg.activate(".")         # activate summer school environment
Pkg.instantiate()         # install Package version from manifest.toml
Pkg.status()
```

### Python via Conda {-}
```{r}
library(reticulate)
## create conda environment if needed
if(!("SummerSchoolTest" %in% conda_list()[,1])) {
    conda_create(environment = "environment.yml")
}
use_condaenv("SummerSchoolTest") # activate environment
conda_python(envname = "SummerSchoolTest") # show the active python binary
```


# Rendering Rmarkdown files

`Rmarkdown` files (typical file eding is `*.rmd`) are compiles with
the R packages `rmarkdown`. Additionally `pandoc` may need to be
installed.

Then simply run:
```{r, eval=FALSE}
library(rmarkdown)
rmarkdown::render("example.rmd")
```

In Rstudio there is a button to execute this command.

# Show and evaluate code

We can show the code and the result:
```{r}
x <- rnorm(10)^2
summary(x)
```

Or just run the code but don't show it, only the result:
```{r, echo=FALSE}
y <- sin(x)
mean(y)
```

Or just run and show the code but hide the result
```{r, results='hide'}
temp.result <- cos(x)
```

Or only show the code but do not run it at all:
```{r, eval=FALSE}
this.will.error <- 11 + "22"
```

Like equations, we can also inline code. For example, a circle with
the radius `r mean(x)` has an area of `r pi * mean(x)^2`.



## Use tabs to for different languages

We use tabs to organize the solution in different languages.


### Solution {.tabset}

#### R {-}
This the R solution:
```{r}
x <- rnorm(10)^2
mean(x)
```

```{r}
hist(y)
```

#### Julia {-}
Needs the R package `JuliaCall` installed!
```{julia, eval=TRUE}
using Plots
α = randn(100);
plot(α, randn(100), label="That's a mess")
```

#### Python {-}
This the Python solution:
```{python}
import numpy as np
x = np.random.normal(10)
np.mean(x)
```

Another code snippet
```{python}
import matplotlib.pyplot as plt
import numpy as np

xpoints = np.array([1, 2, 6, 8])
ypoints = np.array([3, 8, 1, 10])

plt.plot(xpoints, ypoints)
plt.show()
```



# Show and hide solutions

We want to compile two documents, one with and without the solutions.
Set the set the parameter `showsolutions` in the header accordingly
```
params:
  showsolutions: TRUE
```
to `TRUE` or `FALSE`. Alternatively you can also overwrite the parameter during compilation:
```{r, eval=FALSE}
rmarkdown::render("example.rmd", params=list(showsolutions=FALSE))
```

The solutions for every task needs to be written in a separate document that is
optionally included with
````
## ```{r, child = if(params$showsolutions){"solutions.Rmd"}}
## ```
````

<!-- inlcude the solutions if `showsolutions==TRUE` -->
```{r, child = if(params$showsolutions){"exercise_template_solution_1.Rmd"}}
```

# Math

Latex syntax is used to for equation. Either inline $a=\int g(x)\, \text{d}x$ or as
separate equation like this

$$M(x, \theta) = \theta_1 + \theta_2 x^2 + \theta_1 \theta_2 \sqrt{x} .$$


# More on Rmarkdown

[This online book](https://bookdown.org/yihui/rmarkdown/) provides a very comprehensive description what can be done with Rmarkdown.
