---
title: "Hierarchical Modeling with JAGS"
author: "Helen Moor"
output:
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float: true
    css: ../eawag.css
params:
  showsolutions: TRUE
---


**Objectives**

* Brief review of the concept of hierarchical statistical models.
* Example how to use `JAGS` together with R to do Bayesian inference for parameters and states of potentially hierarchical models.
* Build an ecological occupancy model and the dynamic occupancy model of Kéry & Royle (2007) in `JAGS`.

**Prerequisites**

* `JAGS` installed from http://mcmc-jags.sourceforge.net/
* `R` and the `R` library
  [`jagsUI`](https://cran.r-project.org/web/packages/jagsUI/index.html)
  to interface Jags (for Julia there is [`Jags.jl`](https://github.com/JagsJulia/Jags.jl))
* Dataset `BomVarData.csv`



Hierarchical statistical models are sets of coupled probability models
that are conditionally related to each other. As such, hierarchical
models provide a broad framework that encompasses different kinds of
models, from generalized linear mixed models (often called multilevel
models) to Bayesian belief networks to customized process-based models
in ecology, physiology, hydrology, etc.

Hierarchical modeling provides a general and flexible framework for
building complex statistical models from simple building
blocks. `JAGS` is a software that makes it easy to program and fit
complex hierarchical models, because users don't have to explicitly
state the joint probability distribution of the entire model. Instead,
the model is represented by submodels and their conditional
dependencies. This allows for the creative development of complex
models for which no standard packages are available.

`JAGS` uses Gibbs sampling to do Bayesian inference for parameters and
states of (potentially) hierarchical models. Other softwares and
packages can be used for probabilistic programming, such as `Stan`
(interfaced to `R`with package `rstan`) or the python library `PyMC3`
(Krapu & Borsuk 2019). An advantage of `JAGS` over Stan is that `JAGS`
can accommodate discrete variables, such as Bernoulli-distributed
presence-absences.


# 1. Toad colonization 🐸


To demonstrate model development with `JAGS` with use we use field
observations of toads. The dataset used here is a subset of the
long-term monitoring data set of observations of the [yellow-bellied
toad](https://en.wikipedia.org/wiki/Yellow-bellied_toad) (*Bombina
variegata*) in the canton of Aargau in Switzerland.

The core data is detection/non-detection of this species observed at
$N=379$ sites (ponds) $i$ during up to 3 repeat visits $j$, over
multiple years $t$ from 1999 to 2009 ($T=11$). Sites are nested in
four different regions $R$.

We first collapse these data across years and visits to observed
presence-absence per site, and fit a very simple model to estimate the
average expected probability of occurrence (occupancy). We conduct a
naive analysis that ignores detection error.


```{r, echo=T, eval=F}
# read data
dat <- read.csv(file = "../../data/BomVarData.csv")

# number of sites
nsites <- length(unique(dat$SiteID))

# extract the site data only:
sites <- unique(dat[, !(names(dat) %in% c("Visit", "Year", "Detection", "ObsID", "SiteVisited"))])
sites <- sites[(order(sites$SiteID)),]

# extract detection/nondetection data per site, pooling all visits and years:
# was the toad ever seen in site i?
seen_i <- aggregate(dat$Detection, by = list(SiteID=dat$SiteID), FUN=max, na.rm=TRUE)

# number of regions
nregions <- length(unique(sites$Region))

# Region index: in JAGS we need to use integer indices to encode
#  categorical data; indices must start at 1 and be consecutive
Region <- as.integer(as.factor(sites$Region))
```


# 2. Presence/absence model

We describe the presence/absence of the toads by an generalized linear
mixed model (GLMM), a relatively simple hierarchical model. It allows
to account for differences between regions in the expected occurrence
probability $\theta$.


\begin{align*}
y_{i} & \sim \text{Bernoulli}(\theta_{i}) \\
\theta_{i} & = \text{logit}^{-1}(\alpha_{R(i)}) \\
\alpha_{R(i)} & \sim \text{Normal}(\mu, \tau) \\

\mu & \sim \text{Normal}(0, 0.001)\\
\tau & \sim \text{Gamma}(0.5, 0.5)
\end{align*}

where $\text{logit}^{-1}(x) =  \exp(x)/(1 + \exp(x))$.

<!-- We here specify the intercept of occurrence probability as a
random variable with its own model; in other words, we have added a
random effect for region. -->
Regional occurrence probabilities are
generated by a common process with expectation $\mu$ and _precision_
$\tau = 1 /\sigma^2$, called hyperparameters with their own prior
distributions (hyperpriors).

### Implementation in `JAGS`

A `JAGS` model must be defined in a separate text file; we can writ
this file from R or use any text editor. The `JAGS` model description goes inside curly
brackets, preceded with the word model:`model { }`. Use hashtags to
comment code. The model is specified as a series of relations between
variables. Every variable must be either defined in the model or given
as data. Relations can be deterministic (`<- `) or stochastic (`~
`). `JAGS` automatically detects from the data provided which
variables are observed and which ones have to be inferred. The order
in which the model components are specified does not matter, it gets
compiled into a DAG before running. `JAGS` then does Gibbs sampling
from the conditional distributions.

Steps:

1. write the `JAGS` model to a file
    + Note: in `JAGS`, levels of categorical variables are specified as integers, that must be consecutive and start at 1
    + Note: `JAGS` specifies normal distributions by the mean and precision $\tau = 1/\sigma^2$
2. run the model from `R`
    + bundle input data in a list  `JAGS`
    + define initial values (optional)
    + list names of parameters to be monitored
    + run the model using function `jags()` from package `jagsUI`


We save the following JAGS code in a file called `glmm_v1.jags`. Note
how closely the syntax follows the  mathematical equations.
```{r, echo=TRUE, eval=FALSE}
model{

    ## Is site i occupied?
    for(i in 1:nsites){
        y[i] ~ dbern(theta[i])
        logit(theta[i]) <- alpha_R[Region[i]]
    }

    ## Priors:
    for(r in 1:nregions){
        alpha_R[r] ~ dnorm(mu,tau)
    }

    ## Hyperpriors:
    mu ~ dnorm(0, 0.001) # corresponds to a Normal distribution with variance = 1/0.001
    tau ~ dgamma(0.5, 0.5) # gamma prior on precision (cf. Gelman 2006b)
}
```

```{r, echo=TRUE, warning=FALSE, eval=F}
library(jagsUI)

### Run the JAGS model from R:---------------------------
### Bundle data
in.data <- list(y = seen_i$x,
                nsites = nsites,
                Region = Region,
                nregions = nregions)

### Initial values
### -> can be omitted; if no inits are given, JAGS picks them automatically

### Parameters to monitor
params <- c("alpha_R", "mu", "tau") # "sd"

### Run the model
# MCMC settings
ni <- 3000  # iterations
nt <- 10    # thinning
nb <- 1000  # burn-in
nc <- 3     # chains


out.v1 <- jags(data = in.data,
                   parameters.to.save = params,
                   parallel = T,
                   model.file=paste0("./JAGS_models/glmm_v1.jags"),
                   n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb,
                   verbose=FALSE, store.data = F)

out.v1

# regional estimates:
plogis(out.v1$mean$alpha_R)  # plogis(x) = exp(x)/(1 + exp(x))
# grand mean:
plogis(out.v1$mean$mu)
```

### Tasks

Extend the `JAGS` model from above to include the effect of pond size on occupancy.

Steps in extending a model:

* `JAGS`:
  + write an updated version of the process or observation model
  + add priors for any new parameters
* `R`:
  + update input data
  + update inits (optional)
  + update names of parameters that are monitored



```{r, child = if(params$showsolutions){"exe_4_solution_2.Rmd"}}
```

# 3. Static occupancy model

Occupancy models account for the observation process as separate from
the true, latent ecological process. They do this by introducing a new
random variable $z$, which represents the true latent, but imperfectly
observed occupancy state (MacKenzie et al. 2002).
Detection/non-detection data from repeat visits to the same sites are
exploited to estimate detection probability. At least some sites must
have been visited repeatedly, during a period of closure (i.e., during
a period within which the true state is assumed to remain constant).

Detection/non-detection data $y_{ij}$ at site $i$ during repeat
surveys $j$ are then described by a hierarchical model that contains
one submodel for the only partially observed true state $z$ (occupied
or not occupied), and another submodel for the actual observations
$y$.  The actual observations result from both the particular
realization of the ecological process and of the observation process.


The observation process yields observations for each site $i$ during
visit $j$, *conditional on* the true state $z_{i}$ and detection
probability $p$

$$
y_{ij} \sim \text{Bernoulli}(z_{i} p)
$$
Detection probability $p$ is here assumed to be constant across sites
and visits. We can easily formulate the model so as to let detection
probability vary across sites and/ or visits (i.e., $p_{ij}$).


The ecological process yields the true, latent state $z_{i}$ of site
$i$ (presence/ absence of a species)
$$
z_{i} \sim \text{Bernoulli}(\psi)
$$
with occupancy probability $\psi$ .

The model is completed with (flat) priors $\psi \sim \text{Uniform}(0,1)$ and $p \sim \text{Uniform}(0,1)$

In `JAGS`, the components of this hierarchical model are encoded in
the same way as the above algebraic statements, omitting the need to
explicitly formulate the joint likelihood.


## Data

We will now use the detection/non-detection data of this species
during up to 3 repeat visits $j$ in each site $i$, over multiple years
$t$ from 1999 to 2009 ($T=11$), along with information on the sites
and the observation process (the identity of the field worker/
observer).

The observation data is arranged in an array:

*  `y_ijt`: Array of observations (detection/ non-detection) in sites $i$, repeat visit $j$, year $t$

Covariate data, ecological model:

* `sites`: Dataframe containing potentially informative environmental variables describing the sites; we are especially interested in the water surface area (or size) of the ponds and the effect of water table fluctuations. These are features that can be controlled when constructing ponds.

Covariate data, observation model:

* `obs_ijt`: Array with identity of observer, dimensions [site $i$, visit $j$, year $t$]. Observer identity is encoded as an integer (1:134 over entire data set).

<!-- Helper data: -->

<!-- * `visited_ijt`: Array specifying whether a site was visited during visit $j$ (1) or not (0), dimensions  [site $i$, visit $j$, year $t$]. -->

**Important** when formatting data:

* `JAGS` loops over all dimensions of vectors, matrices, etc. Good bookkeeping is essential! We must ensure that the data are all in the right order, e.g. that the order of sites is the same in x (`sites$Fluctuations`) and in y. It is a good idea to always label data, e.g. use dimnames for arrays.
* Categorical variables are encoded in `JAGS` as integer indices; they must be consecutive and start at 1 (_not at zero_).
* Where possible, standardize covariates (to mean zero and variance = 1), in order to improve convergence.

```{r, echo=T, eval=F}
nsites <- length(unique(dat$SiteID))
nyears <- length(unique(dat$Year))
nvisits <- length(unique(dat$Visit))

# extract the site data only:
sites <- unique(dat[, !(names(dat) %in% c("Visit", "Year", "Detection", "ObsID", "SiteVisited"))])
sites <- sites[(order(sites$SiteID)),]

# extract the detection/nondetection data and store in array
# with dimensions [sites, visits, years]
y_ijt <- array(data = dat$Detection,
               dim = c(nsites, nvisits, nyears),
               dimnames=list(sites$SiteID, c(1:3), 1999:2009))

dim(y_ijt) # sites, visits, years
dimnames(y_ijt)[[3]]


# extract the observer identity and store in array with dimensions [sites, visits, years]
obs_ijt <-  array(data = dat$ObsID,
                  dim = c(nsites, nvisits, nyears),
                  dimnames=list(sites$SiteID, c(1:3), 1999:2009))
nobservers <- length(unique(dat$ObsID))
```



## Occupany model for a single year

We build the occupancy model for the first year, which will be needed
to provide initial conditions for the dynamic model. We subset our
data to the first year, and formulate the model for a single year only
(note the indices!). We start with above model that accounts for
regional differences in occupancy $\psi_R$:

$$
z_{i,t=1} \sim \text{Bernoulli}(\psi_{R,t=1})
$$
$$
y_{ij,t=1} \sim \text{Bernoulli}(z_{i,t=1} p)
$$

While the indexing may seem tedious, it makes explicit over which
dimensions the random variables or parameters vary. To begin with,
we here assume detection probability to be constant across sites and
visits.

We save the JAGS models as `occ_v0.jags`:
```{r, echo=T, warning=FALSE, eval=FALSE}
model{
    ## Ecological process model

    for(i in 1:nsites){
          z[i] ~ dbern(psi1[i])
          logit(psi1[i]) <- alpha.psi1[Region[i]]
    }

    ## Observation model:
    for(i in 1:nsites){
          for (j in 1:nvisits){
            y[i,j] ~ dbern(muy[i,j])
            muy[i,j] <- z[i] * p
          }
      }

    ## Priors:

    ## Ecological model
    for(r in 1:nregions){
        alpha.psi1[r] ~ dnorm(mu, tau)
    }
    ## Hyperpriors
    mu ~ dnorm(0,0.001)
    tau ~ dgamma(0.5, 0.5) # gamma prior on precision (cf. Gelman 2006b)

    ## Observation model
    p ~ dunif(0,1)    # (constant) detection probability
}
```

We compile all needed data in R as list and then run the JAGS model
```{r, echo=T, eval=F}
### Bundle data (we here use data from the first year only)
in.data <- list(y = y_ijt[,,1],
                nsites = nsites,
                nvisits = nvisits,
                nregions= nregions,
                Region = Region
                )

### Initial values
### We use observed occurrences as inits for z
temp <- y_ijt; temp[is.na(temp)] <- 0  # remove NAs to avoid conflicts
z.inits <-apply(temp, c(1,3), max); rm(temp)

inits <- function() list(z = z.inits[,1],
                         p = runif(n = 1, 0, 1))

### Parameters to monitor
params <- c("p",
            "alpha.psi1")

### Run the model
# MCMC settings
ni <- 5000  # iterations
nt <- 10    # thinning
nb <- 1000  # burn-in
nc <- 3     # chains


out.v0 <- jags(data = in.data,
               inits = inits,
               parameters.to.save = params,
               parallel = TRUE,
               model.file=paste0("./JAGS_models/occ_v0.jags"),
               n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb,
               verbose=FALSE, store.data = TRUE)

```


## Tasks 1

Extend the ecological process model to test whether the occurrence of
this toad depends on the size (surface area) of the pond. Check
whether there is a unimodal (quadratic) instead of linear response of
occupancy probability to pond area.


$$
z_{i,t=1} \sim \text{Bernoulli}(\psi_{i,t=1}),
$$
$$
logit(\psi_{i,t=1}) = \alpha_{\psi1, R(i)} + \beta_{area} Area_{i},
$$

Additionally, try what happens when there are NAs in the explanatory variable!




```{r, child = if(params$showsolutions){"exe_4_solution_3.1.Rmd"}}
```

## Task 2

Second, we want to improve the observation model, by accounting for variation in detection probability between observers. Observers are trained volunteers with different levels of experience. The identity of the observer may vary between sites, visits, and years: detection probability becomes $p_{ijt}$ and we need to loop accordingly. We estimate detection probability with a random effect of observer identity, where observers vary around a common mean $\mu_{obs}$:

$$
logit(p_{ij,t=1}) = \alpha_{obs} + \beta_{obs(ij,t=1)},   \\   \beta_{obs} \sim \text{Normal}(0,\sigma_{obs}^2).
$$
*NOTE*: indices for categories or factor levels in `JAGS` are encoded as *consecutive* integers, starting from 1.


```{r, child = if(params$showsolutions){"exe_4_solution_3.2.Rmd"}}
```


# 4. Dynamic occupancy model

The dynamic occupancy model extends the static, single-year model to
the dynamic change in occupancy over time. It is based on
metapopulation theory and describes how the true state (occupied or
empty) of each site evolves over time as a function of two processes:
colonization and persistence.

We initiate the dynamic model with a simple occupancy model for the
occupancy probability in the first year. This provides the starting
conditions $z_{i, t=1}$ for the dynamic process.

$$
z_{i,t=1} \sim \text{Bernoulli}(\psi_{i,t=1})
$$

After the first year, the true state of each site depends only on its
state in the previous year and the probabilities of colonization and
persistence. Empty patches can get colonized in the next time step
with probability $\gamma$ or remain empty with probability $1-
\gamma$.  Occupied patches can either persist (i.e., remain occupied)
with probability $\phi$ or go locally "extinct" with probability $1-
\phi$. The realization of this process over years $t = 2, ...,T$ can
be formulated for each site $i$ and year $t$ as conditional on the
true state in the previous time step and the two process
probabilities:


$$
z_{i, t} \sim \text{Bernoulli}( (1-z_{i, t-1}) \gamma_{it} + z_{i, t-1} \phi_{it} )
$$

This has the advantage that we can learn something about the dynamic
process rates, not only the resulting occupancy probability. Finally,
observations (detection/ non-detection data) are again related to the
true, latent occupancy state via the observation model:


$$
y_{ijt} \sim \text{Bernoulli}(z_{it} p_{ijt})
$$
Detection probability $p_{ijt}$ is here allowed to vary between sites, visits, and years.


The aim of the dynamic occupancy model is to estimate the
probabilities of colonization and persistence, and their dependence on
site characteristics, using detection/ non-detection data and
accounting for observation uncertainty.


Update the `JAGS` model to use all data and include the dynamic
process. Start by assuming constant colonization and persistence
probabilities.

Covariate effects on colonization and persistence probabilities can
easily be included using GLMs. Note that these explanatory models
don't have to be linear models, but can take any functional form.

We save the JAGS model in `dynocc_v1.jags`.
```{r, echo=T, eval=F, warning=FALSE}
model{
    # Ecological process model:
    for(i in 1:nsites){
        # initial occupany in year 1
        z[i,1] ~ dbern(psi1[i])
        logit(psi1[i]) <- alpha.psi1[Region[i]]

        # dynamic change year 2:T
        for (k in 1:(nyear-1)){
            z[i,(k+1)] ~ dbern(muZ[i,(k+1)])
            muZ[i,(k+1)]<- ((1-z[i,k])*gamma[i,k]) + (z[i,k]*phi[i,k])

            logit(gamma[i,k]) <- alpha.gamma
	        logit(phi[i,k]) <- alpha.phi
        }
    }


    # Observation model:
    for(i in 1:nsites){
          for (j in 1:nvisits){
            for(k in 1:nyear){
              y[i,j,k] ~ dbern(muy[i,j,k])
              muy[i,j,k] <- z[i,k] * p[i,j,k]
              logit(p[i,j,k]) <- alpha.obs + beta.obs[obs[i,j,k]]
            }
         }
      }


    # Priors:

    ## Ecological process model:
    # Occupancy year 1:
    for(r in 1:nregions){
       alpha.psi1[r] ~ dnorm(mu, tau)
    }
    # Hyperpriors:
    mu ~ dnorm(0,0.001)
    tau ~ dgamma(0.5, 0.5)

    # Colonization gamma, persistence phi:
    alpha.gamma ~ dnorm(0, 0.001)
    alpha.phi ~ dnorm(0, 0.001)

    ## Observation model:
     alpha.obs ~ dnorm(0, 0.001)
    # random observer effects:
    for (obs in 1:nobservers){
      beta.obs[obs] ~ dnorm(0, tau.obs)
    }
    tau.obs <- 1/ pow(sd.obs,2) # tau = 1/variance
    sd.obs ~ dunif(0,10)
}
```



Again, we run this model from R:
```{r, echo=F, eval=F, warning=FALSE}
### 2. UPDATE INDATA, INITS, PARAMETERS
### Data prep:
# pass entire time series

# how many observers are there across all years?
nobservers <- length(unique(c(obs_ijt)))

### Bundle data (first year only)
in.data <- list(y = y_ijt,
                nsites = nsites,
                nvisits = nvisits,
                nregions = nregions,
                Region = Region,
                obs = obs_ijt, # observer identities in all years
                nobservers = nobservers, # number of unique observers (all years)
                nyear = nyears # number of years
                )

### Initial values
### use observed occurrences as inits for z
temp <- y_ijt; temp[is.na(temp)] <- 0  # remove NAs to avoid conflicts
z.inits <-apply(temp, c(1,3), max)

inits <- function() list(z = z.inits,
                          alpha.psi1 = runif(n=nregions, -3, 3),
                         alpha.obs = runif(n=1, 0.1, 1),
                         sd.obs = runif(n = 1, 0.5,3))

### Parameters to monitor
params <- c("alpha.psi1",  # parameters for occupancy probability in year 1
            #"mu", "tau",  # hyperparameters for regional differences psi1
            "alpha.gamma", "alpha.phi", # colonization and persistence intercepts
            "alpha.obs", "sd.obs" ) #  observer effect
            #"beta.obs") # p estimates for individual observers

### Run the model
# MCMC settings  (ca. 3 mins)
ni <- 8000  # iterations
nt <- 10    # thinning
nb <- 3000   # burn-in
nc <- 3     # chains

out.dyn1 <- jags(data = in.data,
            inits = inits,
            parameters.to.save = params,
            parallel = TRUE,
            model.file=paste0("./JAGS_models/dynocc_v1.jags"),
            n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb,
            verbose=TRUE, store.data = TRUE)
```

Then we look at the results
```{r, echo=F, eval=F, warning=FALSE}
### inspect the outcome:
out.dyn1

plot(out.dyn1) # traceplots and densities

## What have we learnt about colonization and extinction probabilities?
## out.dyn1$sims.list contains the samples from posteriors
str(out.dyn1$sims.list)

par(mfrow=c(1,2))
## plot colonization intercept, on probability scale:
hist(plogis(out.dyn1$sims.list$alpha.gamma), main="Colonization probability")
abline(v=plogis(out.dyn1$mean$alpha.gamma), col="red")
#hist(plogis(out.dyn1$sims.list$alpha.gamma))

## plot persistence intercept, on probability scale:
hist(plogis(out.dyn1$sims.list$alpha.phi), main="Persistence probability")
abline(v=plogis(out.dyn1$mean$alpha.phi), col="red")
```


## Tasks

### Extend the dynamic occupancy model

Test for effects of covariates on colonization and persistence probabilities:

* Does colonization probability depend on
  + the size, i.e. area, of the pond? Linearly or unimodally?
  + whether the water table fluctuates?
  + (whether the pond is a newly constructed pond?)

* Does persistence probability depend on
  + the size, i.e. area, of the pond? Linearly or unimodally?
  + whether the water table fluctuates?
  + ...

We can additionally let `JAGS` calculate derived variables, such as a summary of the actual number of sites occupied. We will obtain full posteriors for such quantities as well.

* Add a derived quantity: what is the estimated true number of occupied ponds in each year?


```{r, child = if(params$showsolutions){"exe_4_solution_4.Rmd"}}
```


# 5. Summary

* `JAGS` allows you to do inference for complex hierarchical models.
* Specify a `JAGS` model within `model{ }` as a series of stochastic (`~`) and deterministic (`<-`) relations.
* Run `JAGS` from `R`:
  + edit and bundle required data (be careful with dimensions and order; consecutive integer indices for levels of categorical variables)
  + specify initial values to help convergence (else let `JAGS` choose automatically from the prior)
  + specify parameters to be monitored
  + set MCMC settings and run the model
* Working with `JAGS` output:
  + the `jagsUI` model object contains a lot of helpful information; storing the used data and model aids reproducibility
  + working with the full joint posteriors (`out$sims.list`) is the most flexible


# References

Gelman, A. (2006). Multilevel (hierarchical) modeling: What it can and cannot do. Technometrics, 48(3), 432–435. https://doi.org/10.1198/004017005000000661

Gelman, A. (2006b). Prior distributions for variance parameters in hierarchical models (Comment on Article by Browne and Draper). Bayesian Analysis, 1(3), 515–534. https://doi.org/10.1214/06-BA117A

Krapu, C., & Borsuk, M. (2019). Probabilistic programming: A review for environmental modellers. Environmental Modelling and Software, 114(January), 40–48. https://doi.org/10.1016/j.envsoft.2019.01.014

MacKenzie, D. I., Nichols, J. D., Lachman, G. B., Droege, S., Royle, A. A., & Langtimm, C. A. (2002). Estimating site occupancy rates when detection probabilities are less than one. Ecology, 83(8), 2248–2255. https://doi.org/10.1890/0012-9658(2002)083[2248:ESORWD]2.0.CO;2

Royle, J. A., & Kéry, M. (2007). A Bayesian state-space formulation of dynamic occupancy models. Ecology, 88(7), 1813–1823. https://doi.org/10.1890/06-0669.1
